Dissecting the causal question underlying the association between cancer and ADRD: an opportunity to understand sources of bias

### 2021-07-15 04_reports/manuscript.Rmd
cleaning

Many observational studies have consistently found that individuals with cancer have a lower risk of developing of Alzheimer´s disease or related dementias (ADRD) when compared to individuals with no history of cancer [@ma2014;@hanson2016;@vanderwillik2018;@ospina2020]. These findings have motivated substantial research toward mechanistic explanations, including molecular and genetic pathways that may explain this association[@behrens2009;@harris2014;@nudelman2019;@Papin2020;@driverbiogeront2014;@olson2019;@li2021;@driverpin12015]. These research inquiries inevitably lead to discussions of repurposing or augmenting current cancer chemotherapeutics for ADRD [@Snyder2017].

Nevertheless, inferring any treatment or mechanistic effects from the observed cancer-ADRD inverse association is not straightforward. Researchers have raised concerns related to the competing event of death, unmeasured confounding, and ascertainment error that could explain these results[@driverbiogeront2014;@ganguli2015]. However, understanding these or other sources of bias first requires making explicit the causal question. Moreover, this is one step toward tying a research study to a question that is relevant to decision-making[@didelez2016;@labrecque2017].

@@

To explore how we might learn about this effect using real-world data on cancer and ADRD, we progressively build a causal directed acyclic graph to connect this particular causal question to the observable data and the assumptions we rely on to study the effect. We exemplify these challenges and how they translate into the analytic decision using data collected from the Rotterdam Study, a population-based cohort study.

@@

In the observational setting, confounding could explain the observed association between Pin1 and ADRD. In the directed acyclic graph [@whatif2020] of Figure 1, we present Pin1 expression as $P_{t-1}$ and ADRD diagnosis over time as $Y_{t+1}$. Both nodes may share causes $L$, and to assess the causal relationship between $P_{t-1}$ and $Y_{t+1}$ we would require adjusting for $L$. Previous studies have described age, sex, educational level and race/ethnicity as the minimal adjusting set of covariates[@ospina2020]. However, environmental and behavioral factors such as smoking may translate into Pin1 over-expression and are also related to the development of ADRD[ref].

```{tikz, echo = FALSE, fig.cap = "Pin1 as biomarker and risk of dementia"}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm]
\node (p) {$P_{t-1}$};
\node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L$};
\node [gray!30, below of = p, yshift = 1cm] (pstar) {$P^*_{t-1}$};
\node [gray!30, right of = p] (dp) {$D_t$};
\node [gray!30, right of = dp] (p2) {$P_t$};
\node [gray!30, below of = p2, yshift = 1cm] (pstar2) {$P^*_t$};
\node [gray!30, right of = p2] (dp2) {$D_{t+1}$};
\node [right of = dp2] (y) {$Y_{t+1}$};
\node [gray!30, below of = dp2, yshift = 1cm, xshift = +1cm](cd) {$C$};
\draw[->, gray!30] (p) to (pstar);
\draw[->, gray!30] (p2) to (pstar2);
\draw[->] (l) -- (p);
\draw[->, gray!30] (l) -- (p2);
\draw[->, gray!30] (l) to (dp);
\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (dp2);
\draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
\draw[->, gray!30] (pstar) -- (dp);
\draw[->, gray!30] (pstar2) -- (dp2);
\draw[->, gray!30] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
\draw[->, gray!30] (dp) -- (p2);
\draw[->, gray!30] (dp2) -- (y);
\draw[->, gray!30] (cd) -- (dp2);
\draw[->, gray!30] (cd) -- (y);
\draw[->, gray!30, semithick] (p2) to [out=+45,in=+135, looseness=0.3] (y);
\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
\end{tikzpicture}
```

Currently, Pin1 expression is not an available biomarker for population-based research, so at best we can only rely on a proxy for it. Because Pin1 over-expression is present in tumors, and tumors are only measured through diagnosis, previous studies have considered Pin1 over-expression as the explanation behind the inverse association between cancer diagnosis and ADRD, though Pin1 was not explicitly part of the research question[@driver2012;@musicco2013;@freedman2016;@bowles2017;@frain2017;@schmidt2017;@sun2020;@ording2020]. Unlike measuring Pin1 at the same time to all participants (thought this would not necessarily mean this would be the ideal time to measure it, we discuss this point further in the discussion section), cancer diagnosis is a collected over time. We depict this feature in Figure 2, where $P^*_{t-1}$ and $P^*_t$ represent _cancer diagnosis_ over time, the measured proxy of $P_{t-1}$ and $P_t$ respectively. Although this means we would measure the association between cancer diagnosis over time and ADRD in the observed data, we are assuming that the captured effect is only through the pathway that involves Pin1 expression over time.

```{tikz, echo = FALSE, fig.cap = "Cancer diagnosis as proxy for Pin1 expression"}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm]
\node (p) {$P_{t-1}$};
\node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L$};
\node [below of = p, yshift = 1cm] (pstar) {$P^*_{t-1}$};
\node [gray!30, right of = p] (dp) {$D_t$};
\node [right of = dp] (p2) {$P_t$};
\node [below of = p2, yshift = 1cm] (pstar2) {$P^*_t$};
\node [gray!30, right of = p2] (dp2) {$D_{t+1}$};
\node [right of = dp2] (y) {$Y_{t+1}$};
\node [gray!30, below of = dp2, yshift = 1cm, xshift = +1cm](cd) {$C$};
\draw[->] (p) to (pstar);
\draw[->] (p2) to (pstar2);
\draw[->] (l) -- (p);
\draw[->] (l) -- (p2);
\draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
\draw[->, gray!30] (l) to (dp);
\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (dp2);
\draw[->, gray!30] (dp) -- (p2);
\draw[->, gray!30] (pstar) -- (dp);
\draw[->, gray!30] (pstar2) -- (dp2);
\draw[->, gray!30] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
\draw[->, gray!30] (dp2) -- (y);
\draw[->, gray!30] (cd) -- (dp2);
\draw[->, gray!30] (cd) -- (y);
\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
\draw[->, semithick] (p2) to [out=+45,in=+135, looseness=0.3] (y);
\end{tikzpicture}
```

A challenge that arises by choosing cancer diagnosis as the proxy for Pin1 is defining the time zero, the time where eligibility criteria is met, where exposure is measured and the time where the screening for ADRD would begin after having the exposure measured[@hernanInmortal2016]. The eligibility criteria to participate in a trial for Pin-targeting drug will not necessarily share the same criteria to be recruited for the cohort setting. The latter may not align with the time cancer diagnosis is measured since it happens over follow-up and this situation may introduce inmortal-time bias[@hernanInmortal2016]. For example, a study performed using data from the Framingham study [@driver2012] defined the exposed group with cancer as those participants with prevalent or incident cancer diagnosis (alternatively defined as "ever cancer" [@hanson2016]). This meant that a participant who had cancer diagnosis over follow-up contributed all their person-time to the cancer arm, including the time prior to the cancer diagnosis. By defining the exposure after the time of inclusion to the cohort, only participants who remain alive and free of ADRD diagnosis over follow-up can be defined as "ever cancer" [@anderson1983, @hernanInmortal2016].

This problem is avoided by recognizing the time-varying nature of cancer diagnosis. Several studies have considered cancer diagnosis as a time-dependent exposure. In this way, the time prior to cancer diagnosis is allocated to the non-exposed arm, and the time after cancer diagnosis to the exposed arm [@white2013; @hanson2016; @bowles2017]. The price we pay with this approximation is that implicitly, this means that Pin1 would over-express at the time of cancer diagnosis and not before, which is biologically implausible. Moreover, cancer diagnosis will only be measured in the subset of participants who are alive over follow-up. Thus, we included death prior to cancer diagnosis as $D_t$ and an arrow between $D_t$ and $P_t$ that represents a deterministic association such as that $P_t$ is only observed if $D_t$ is zero in Figure 3. In addition, we added an arrow between $L$ and $D_t$, since covariates such as smoking may affect Pin1 over-expression but also affect the risk of death due to other causes such as from chronic obstructive pulmonary disease. Although $L$ is time-varying in nature, we only depict $L$ at one time-point for readability.

```{tikz, echo = FALSE, fig.cap = "Time-varying cancer diagnosis and inmortal-time bias"}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm]
\node (p) {$P_{t-1}$};
\node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L$};
\node [below of = p, yshift = 1cm] (pstar) {$P^*_{t-1}$};
\node [right of = p] (dp) {$D_t$};
\node [right of = dp] (p2) {$P_t$};
\node [below of = p2, yshift = 1cm] (pstar2) {$P^*_t$};
\node [gray!30, right of = p2] (dp2) {$D_{t+1}$};
\node [right of = dp2] (y) {$Y_{t+1}$};
\node [gray!30, below of = dp2, yshift = 1cm, xshift = +1cm](cd) {$C$};
\draw[->] (p) to (pstar);
\draw[->] (p2) to (pstar2);
\draw[->] (l) -- (p);
\draw[->] (l) -- (p2);
\draw[->] (l) to (dp);
\draw[->] (dp) -- (p2);
\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (dp2);
\draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
\draw[->, gray!30] (pstar) -- (dp);
\draw[->, gray!30] (pstar2) -- (dp2);
\draw[->, gray!30] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
\draw[->, gray!30] (dp2) -- (y);
\draw[->, gray!30] (cd) -- (dp2);
\draw[->, gray!30] (cd) -- (y);
\draw[->, semithick] (p2) to [out=+45,in=+135, looseness=0.3] (y);
\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
\end{tikzpicture}
```

We now move to the challenge of death as a competing event for ADRD, represented in Figure 4. For interpretability we exclude the time-varying process of cancer diagnosis and focus on Pin1 (and cancer diagnosis) as it had been measured in all participants at time $t$. Given that some participants may die prior to ADRD diagnosis, we can only measure ADRD over follow-up in the individuals who survive long enough to have an ADRD diagnosis. In the causal diagram of Figure 4 we include a node for death after the exposure $P^*_t$ has been measured, represented as $D_{t+1}$ and which acts as a competing event of $Y_{t+1}$ because if a participant dies by $t + 1$, the participant cannot subsequently develop dementia. Furthermore, since $D_{t+1}$ and $Y_{t+1}$ are events related to aging and its consequences, $C$ represents the shared causes of both events such as cardiovascular conditions. We also include an arrow between $L$ and $D_{t+1}$ following the argument discussed previously.

In the setting where $P_t$ represented the targeted-drug for Pin1, and if this drug had no systemic beneficial or harmful side-effects such as that there is no arrow between $P_t$ and $D_{t+1}$, a total effect would quantify the effect of $P_t$ on $Y_{t+1}$ that does not include any pathway mediated through $D_{t+1}$[@young2020]. However, in the context of cancer diagnosis as the proxy for Pin1 over-expression, we cannot rule out the effect of cancer diagnosis on death, represented as the arrow between $P^*_t$ and  $D_{t+1}$. In this setting, a total effect of $P_t$ in $Y_{t+1}$ would include the causal pathway mediated by the effect of cancer diagnosis in mortality, which may translate into an inverse association[@young2020].

To isolate the direct effect of $P_t$ in $Y_{t+1}$ through measurement of $P^*_t$ we have at least two alternatives of causal questions we can ask. We can either imagine a causal question where we decompose the effect of cancer by the different mechanisms that affect dementia and death separately[@stensrud2020]. Alternatively, we can define an scenario where death could have been prevented. The latter is defined as the controlled direct effect, where death is considered as a censoring event and it relies on the assumption that we have measured all $C$ to block the pathway $Y_{t+1} \leftarrow C  \rightarrow D_{t+1} \rightarrow P^*_t \rightarrow P_t$. Previous studies have defined death as a censoring event [@frain2017], although failed to explicitly define how to comply with the independent censoring assumption. Moreover, adjusting for confounders for dementia (such as only adjusting for $L$) will be insufficient to block the pathway mediated by death, time-varying covariates that represent $C$, such as cardiovascular conditions should be considered.

```{tikz, echo = FALSE, fig.cap = "Death as a competing event"}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm]
\node [gray!30](p) {$P_{t-1}$};
\node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L$};
\node [gray!30, below of = p, yshift = 1cm] (pstar) {$P^*_{t-1}$};
\node [gray!30, right of = p] (dp) {$D_t$};
\node [right of = dp] (p2) {$P_t$};
\node [below of = p2, yshift = 1cm] (pstar2) {$P^*_t$};
\node [right of = p2] (dp2) {$D_{t+1}$};
\node [right of = dp2] (y) {$Y_{t+1}$};
\node [below of = dp2, yshift = 1cm, xshift = +1cm](cd) {$C$};
\draw[->, gray!30] (p) to (pstar);
\draw[->, gray!30] (l) -- (p);
\draw[->, gray!30] (l) to (dp);
\draw[->, gray!30] (pstar) -- (dp);
\draw[->, gray!30] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
\draw[->, gray!30] (dp) -- (p2);
\draw[->, semithick, gray!30] (p) to [out=+45,in=+135, looseness=0.3] (y);
\draw[->] (p2) to (pstar2);
\draw[->] (dp2) -- (y);
\draw[->] (cd) -- (dp2);
\draw[->] (cd) -- (y);
\draw[->] (l) -- (p2);
\draw[->] (l) to [out=0,in=90, looseness=0.4] (dp2);
\draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
\draw[->] (pstar2) -- (dp2);
\draw[->, semithick] (p2) to [out=+45,in=+135, looseness=0.3] (y);
\end{tikzpicture}
```

To summarize, the complexity of the causal structure that describes the effect of Pin1 through the proxy of cancer diagnosis in the risk of ADRD illustrate the potential sources of bias, as observed in Figure 5. Even so, this is a simplified version since we omitted other sources of measurement error and the time-varying nature of all nodes and feedback loops between them which would further complicate interpretability[@whatif2020].

```{tikz, echo = FALSE, fig.cap = "Direct effect of Pin1 in the risk of ADRD"}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm]
\node (p) {$P_{t-1}$};
\node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L$};
\node [below of = p, yshift = 1cm] (pstar) {$P^*_{t-1}$};
\node [right of = p] (dp) {$D_t$};
\node [right of = dp] (p2) {$P_t$};
\node [below of = p2, yshift = 1cm] (pstar2) {$P^*_t$};
\node [right of = p2] (dp2) {$D_{t+1}$};
\node [right of = dp2] (y) {$Y_{t+1}$};
\node [below of = dp2, yshift = 1cm, xshift = +1cm](cd) {$C$};
\draw[->] (p) to (pstar);
\draw[->] (l) -- (p);
\draw[->] (l) to (dp);
\draw[->] (pstar) -- (dp);
\draw[->] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
\draw[->] (dp) -- (p2);
\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
\draw[->] (p2) to (pstar2);
\draw[->] (dp2) -- (y);
\draw[->] (cd) -- (dp2);
\draw[->] (cd) -- (y);
\draw[->] (l) -- (p2);
\draw[->] (l) to [out=0,in=90, looseness=0.4] (dp2);
\draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
\draw[->] (pstar2) -- (dp2);
\draw[->, semithick] (p2) to [out=+45,in=+135, looseness=0.3] (y);
\end{tikzpicture}
```

We now turn to an application where we show how these challenges translate into analytic decisions. We will show ways to mitigate or better understand the best of the available data's abilities to inform the possible effect of Pin1 on all cause-dementia.

## 2.2. The Rotterdam Study

We use data collected in the Rotterdam Study, a population-based prospective cohort study among persons living in the Ommoord district in Rotterdam, the Netherlands. Recruitment and initial assessments were held between 1990 and 1993, a second waive of recruitment was held between 2000 and 2001. Participants from first subcohort had follow-up visits between 1993-1995, 1997-1999, 2002-2005, and 2008-2010, second subcohort had follow-up visits between 2004 and 2005, and between 2011 and 2012[@ikram2020]. All participants had data on history of cancer and dementia and incident diagnosis through follow-up until 2015, collected from medical records of general practitioners (including hospital discharge letters) and through linkage with national registries. Date and cause of death was collected on a weekly basis via municipal population registries.

We defined as inclusion criteria being between 60 and 70 years old at study entry, without history of cancer diagnosis, free of cognitive decline or with previous history of dementia. Out of 10998 persons who participated at study entry, `r total_n` were considered eligible. Time to cancer diagnosis, time to dementia diagnosis and death status was measured for all participants. All participants were followed from study entry until dementia diagnosis, death or 20 years after their individual baseline date, whichever occurred first.

>>>>

Many observational studies have consistently found that individuals with cancer have a lower risk of developing Alzheimer´s disease and related dementias (ADRD) when compared to individuals with no history of cancer [@ma2014;@hanson2016;@vanderwillik2018;@ospina2020]. These findings have motivated substantial research toward mechanistic explanations, including molecular and genetic pathways that may explain this association[@behrens2009;@harris2014;@nudelman2019;@Papin2020;@driverbiogeront2014;@olson2019;@li2021;@driverpin12015]. These research inquiries inevitably lead to discussions of repurposing or augmenting current cancer chemotherapeutic for ADRD [@Snyder2017].

Nevertheless, inferring any treatment or mechanistic effects from the observed cancer-ADRD inverse association is not straightforward. Researchers have raised concerns related to the competing event of death, unmeasured confounding, and ascertainment error that could explain these results[@driverbiogeront2014;@ganguli2015]. However, understanding these or other sources of bias first requires making explicit the causal question. Moreover, enumerating an explicit causal question is one step toward tying a research study to a question that is relevant to decision-making[@didelez2016;@labrecque2017].

@@

To explore how we might learn about this effect using real-world data on cancer and ADRD, we progressively build a causal directed acyclic graph (DAG) to connect this particular causal question to the observable data and the assumptions we rely on to study the effect. We exemplify these challenges and how they translate into the analytic decision using data collected from the Rotterdam Study, a population-based cohort study.

@@

In the observational setting, confounding could explain the observed association between Pin1 and ADRD. In the causal diagram [@whatif2020] in Figure 1, we present Pin1 expression as $P_{t}$ and ADRD diagnosis by time $t+k$ as $Y_{t+k}$. Both may share causes $L_1$, and to assess the causal relationship between $P_{t}$ and $Y_{t+k}$ we would thus adjust for $L_1$. Previous ADRD studies have described age, sex, educational level and race/ethnicity as a minimal adjusting set of covariates[@ospina2020]. However, environmental and behavioral factors such as smoking may translate into Pin1 over-expression[@tan2010] and are also related to the development of ADRD[@dementia_lancet].

```{tikz, echo = FALSE, fig.cap = "Pin1 as biomarker and risk of dementia"}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm]
\node (p) {$P_t$};
\node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L_1$};
\node [gray!30, below of = p, yshift = 1cm] (pstar) {$C_t$};
\node [gray!30, right of = p] (dp) {$D_{t+1}$};
\node [gray!30, right of = dp] (p2) {$P_{t+1}$};
\node [gray!30, below of = p2, yshift = 1cm] (pstar2) {$C_{t+1}$};
\node [gray!30, right of = p2] (dp2) {$D_{t+k}$};
\node [right of = dp2] (y) {$Y_{t+k}$};
\node [gray!30, below of = dp2, yshift = 1cm, xshift = +1cm](cd) {$L_2$};
\draw[gray!30, ->] (p) to (pstar);
\draw[gray!30, ->] (l) to (dp);
\draw[gray!30, ->] (pstar) -- (dp);
\draw[gray!30, ->] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
\draw[gray!30, ->] (dp) -- (p2);
\draw[gray!30,->] (p2) to (pstar2);
\draw[gray!30, ->] (dp2) -- (y);
\draw[gray!30, ->] (cd) -- (dp2);
\draw[gray!30, ->] (cd) -- (y);
\draw[gray!30, ->] (l) -- (p2);
\draw[gray!30, ->] (l) to [out=0,in=90, looseness=0.4] (dp2);
\draw[gray!30, ->] (pstar2) -- (dp2);
\draw[gray!30, ->, semithick] (p2) to [out=+45,in=+135, looseness=0.3] (y);
\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
\draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
\draw[->] (l) -- (p);
\end{tikzpicture}
```

Currently, Pin1 expression is not an available biomarker for population-based research, so at best we can only rely on a proxy for it. Because Pin1 over-expression is present in tumors, and tumors are only measured through diagnosis, previous studies have considered Pin1 over-expression as the explanation behind the inverse association between cancer diagnosis and ADRD, though Pin1 was not explicitly part of the research question[@driver2012;@musicco2013;@freedman2016;@bowles2017;@frain2017;@schmidt2017;@sun2020;@ording2020]. Unlike measuring Pin1 at the same time to all participants (thought this would not necessarily mean this would be the ideal time to measure it, we discuss this point further in the discussion section), cancer diagnosis is a collected over time. We depict this feature in Figure 2, where $C_t$ and $C_{t+1}$ represent _cancer diagnosis_ over time, the measured proxy of $P_{t}$ and $P_{t+1}$ respectively. Although this means we would measure the association between cancer diagnosis over time and ADRD in the observed data, we are assuming that the captured effect is only through the pathway that involves Pin1 expression over time. Furthermore, even if we have measurements of $C_{t+1}$ and $Y_{t+k}$ and some subset of $L_1$, our question focused on $P_t$ remains.

```{tikz, echo = FALSE, fig.cap = "Cancer diagnosis as proxy for Pin1 expression"}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm]
\node (p) {$P_t$};
\node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L_1$};
\node [below of = p, yshift = 1cm] (pstar) {$C_t$};
\node [gray!30, right of = p] (dp) {$D_{t+1}$};
\node [right of = dp] (p2) {$P_{t+1}$};
\node [below of = p2, yshift = 1cm] (pstar2) {$C_{t+1}$};
\node [gray!30, right of = p2] (dp2) {$D_{t+k}$};
\node [right of = dp2] (y) {$Y_{t+k}$};
\node [gray!30, below of = dp2, yshift = 1cm, xshift = +1cm](cd) {$L_2$};
\draw[gray!30, ->] (l) to (dp);
\draw[gray!30, ->] (pstar) -- (dp);
\draw[gray!30, ->] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
\draw[gray!30, ->] (dp) -- (p2);
\draw[gray!30, ->] (dp2) -- (y);
\draw[gray!30, ->] (cd) -- (dp2);
\draw[gray!30, ->] (cd) -- (y);
\draw[gray!30, ->] (l) to [out=0,in=90, looseness=0.4] (dp2);
\draw[gray!30, ->] (pstar2) -- (dp2);
\draw[->] (p) to (pstar);
\draw[->] (l) -- (p);
\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
\draw[->] (p2) to (pstar2);
\draw[->] (l) -- (p2);
\draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
\draw[->, semithick] (p2) to [out=+45,in=+135, looseness=0.3] (y);
\end{tikzpicture}
```

A challenge that arises by choosing cancer diagnosis as the proxy for Pin1 is defining the time zero, the time where eligibility criteria is met, where exposure is measured and the time where the screening for ADRD would begin after having the exposure measured[@hernanInmortal2016]. Emulating the eligibility criteria to participate in a trial for Pin-targeting drug will not necessarily be possible in a cohort setting recruiting participants for discordant reasons. The latter may not align with the time cancer diagnosis is measured since it happens over follow-up and this situation may introduce inmortal-time bias[@hernanInmortal2016]. For example, a study performed using data from the Framingham study [@driver2012] defined the exposed group with cancer as those participants with prevalent or incident cancer diagnosis (alternatively defined as "ever cancer" [@hanson2016]). This meant that a participant who had cancer diagnosis over follow-up contributed all their person-time to the cancer arm, including the time prior to the cancer diagnosis. By defining the exposure after the time of inclusion to the cohort, only participants who remain alive and free of ADRD diagnosis over follow-up can be defined as "ever cancer" [@anderson1983, @hernanInmortal2016].

This problem is partly avoided by recognizing the time-varying nature of cancer diagnosis. Several studies have considered cancer diagnosis as a time-dependent exposure. In this way, the time prior to cancer diagnosis is allocated to the non-exposed arm, and the time after cancer diagnosis to the exposed arm [@white2013; @hanson2016; @bowles2017]. The price we pay with this approximation is that implicitly, this means that Pin1 would over-express at the time of cancer diagnosis and not before, which is biologically implausible. Moreover, cancer diagnosis will only be measured in the subset of participants who are alive over follow-up. Thus, we included death prior to cancer diagnosis as $D_{t+1}$ and an arrow between $D_{t+1}$ and $P_{t+1}$ that represents a deterministic association such as that $P_{t+1}$ is only observed if $D_{t+1}$ is zero in Figure 3. In addition, we added an arrow between $L_1$ and $D_{t+1}$, since covariates such as smoking may affect Pin1 over-expression but also affect the risk of death due to other causes such as from chronic obstructive pulmonary disease. Although $L_1$ can be time-varying in nature, we only depict $L_1$ at one time-point for readability.

```{tikz, echo = FALSE, fig.cap = "Time-varying cancer diagnosis and inmortal-time bias"}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm]
\node (p) {$P_t$};
\node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L_1$};
\node [below of = p, yshift = 1cm] (pstar) {$C_t$};
\node [right of = p] (dp) {$D_{t+1}$};
\node [right of = dp] (p2) {$P_{t+1}$};
\node [below of = p2, yshift = 1cm] (pstar2) {$C_{t+1}$};
\node [gray!30, right of = p2] (dp2) {$D_{t+k}$};
\node [right of = dp2] (y) {$Y_{t+k}$};
\node [gray!30, below of = dp2, yshift = 1cm, xshift = +1cm](cd) {$L_2$};
\draw[gray!30, ->] (pstar) -- (dp);
\draw[gray!30, ->] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
\draw[gray!30, ->] (dp2) -- (y);
\draw[gray!30, ->] (cd) -- (dp2);
\draw[gray!30, ->] (cd) -- (y);
\draw[gray!30, ->] (l) to [out=0,in=90, looseness=0.4] (dp2);
\draw[gray!30, ->] (pstar2) -- (dp2);
\draw[->] (p) to (pstar);
\draw[->] (l) -- (p);
\draw[->] (l) to (dp);
\draw[->] (dp) -- (p2);
\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
\draw[->] (p2) to (pstar2);
\draw[->] (l) -- (p2);
\draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
\draw[->, semithick] (p2) to [out=+45,in=+135, looseness=0.3] (y);
\end{tikzpicture}
```

We now move to the challenge of death as a competing event for ADRD, represented in Figure 4. For interpretability we exclude the time-varying process of cancer diagnosis and focus on Pin1 (and cancer diagnosis) as it had been measured in all participants at time $t+1$. Given that some participants may die prior to ADRD diagnosis, we can only measure ADRD over follow-up in the individuals who survive long enough to have an ADRD diagnosis. In the causal diagram of Figure 4 we include a node for death after the exposure $C_{t+1}$ has been measured, represented as $D_{t+k}$ and which acts as a competing event of $Y_{t+k}$ because if a participant dies by $t + k$, the participant cannot subsequently develop dementia. Furthermore, since $D_{t+k}$ and $Y_{t+k}$ are events related to aging and its consequences, $L_2$ represents the shared causes of both events such as cardiovascular conditions. We also include an arrow between $L_1$ and $D_{t+1}$ following the argument discussed previously.

In the setting where $P_{t+1}$ represented the targeted-drug for Pin1, and if this drug had no systemic beneficial or harmful side-effects such as that there is no arrow between $P_{t+1}$ and $D_{t+k}$, a total effect would quantify the effect of $P_{t+1}$ on $Y_{t+k}$ that does not include any pathway mediated through $D_{t+k}$[@young2020]. However, in the context of cancer diagnosis as the proxy for Pin1 over-expression, we cannot rule out the effect of cancer diagnosis on death, represented as the arrow between $C_{t+1}$ and $D_{t+k}$. In this setting, a total effect of $P_{t+1}$ in $Y_{t+k}$ would include the causal pathway mediated by the effect of cancer diagnosis in mortality, which may translate into an inverse association[@young2020].

To isolate the direct effect of $P_{t+1}$ in $Y_{t+k}$ through measurement of $C_{t+1}$ we have at least two alternatives of causal questions we can ask. We can either imagine a causal question where we decompose the effect of cancer by the different mechanisms that affect dementia and death separately[@stensrud2020]. Alternatively, we can define an scenario where death could have been prevented. The latter is defined as the controlled direct effect, where death is considered as a censoring event and it relies on the assumption that we have measured all $L$ to block the pathway $Y_{t+k} \leftarrow L_2  \rightarrow D_{t+k} \rightarrow C_{t+1} \rightarrow P_{t+1}$. Previous studies have defined death as a censoring event [@frain2017], although failed to explicitly define how to comply with the independent censoring assumption. Moreover, adjusting for confounders for dementia (such as only adjusting for $L$) will be insufficient to block the pathway mediated by death, time-varying covariates that represent $L_2$, such as cardiovascular conditions should be considered.

```{tikz, echo = FALSE, fig.cap = "Death as a competing event"}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm]
\node [gray!30](p) {$P_t$};
\node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L_1$};
\node [gray!30, below of = p, yshift = 1cm] (pstar) {$C_t$};
\node [gray!30, right of = p] (dp) {$D_{t+1}$};
\node [right of = dp] (p2) {$P_{t+1}$};
\node [below of = p2, yshift = 1cm] (pstar2) {$C_{t+1}$};
\node [right of = p2] (dp2) {$D_{t+k}$};
\node [right of = dp2] (y) {$Y_{t+k}$};
\node [below of = dp2, yshift = 1cm, xshift = +1cm](cd) {$L_2$};
\draw[gray!30, ->] (p) to (pstar);
\draw[gray!30, ->] (l) -- (p);
\draw[gray!30, ->] (l) to (dp);
\draw[gray!30, ->] (pstar) -- (dp);
\draw[gray!30,->] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
\draw[gray!30, ->] (dp) -- (p2);
\draw[gray!30, ->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
\draw[->] (p2) to (pstar2);
\draw[->] (dp2) -- (y);
\draw[->] (cd) -- (dp2);
\draw[->] (cd) -- (y);
\draw[->] (l) -- (p2);
\draw[->] (l) to [out=0,in=90, looseness=0.4] (dp2);
\draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
\draw[->] (pstar2) -- (dp2);
\draw[->, semithick] (p2) to [out=+45,in=+135, looseness=0.3] (y);
\end{tikzpicture}
```

To summarize, the complexity of the causal structure that describes the effect of Pin1 through the proxy of cancer diagnosis in the risk of ADRD illustrate the potential sources of bias, as observed in Figure 5. Even so, this is a simplified version since we omitted additional arrows from $L_1$ to $C_t$ and $C_{t+2}$ for brevity, as well as other sources of measurement error and the time-varying nature of all nodes and feedback loops between them, which would further complicate interpretability[@whatif2020].

```{tikz, echo = FALSE, fig.cap = "Direct effect of Pin1 in the risk of ADRD"}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm]
\node (p) {$P_t$};
\node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L_1$};
\node [below of = p, yshift = 1cm] (pstar) {$C_t$};
\node [right of = p] (dp) {$D_{t+1}$};
\node [right of = dp] (p2) {$P_{t+1}$};
\node [below of = p2, yshift = 1cm] (pstar2) {$C_{t+1}$};
\node [right of = p2] (dp2) {$D_{t+k}$};
\node [right of = dp2] (y) {$Y_{t+k}$};
\node [below of = dp2, yshift = 1cm, xshift = +1cm](cd) {$L_2$};
\draw[->] (p) to (pstar);
\draw[->] (l) -- (p);
\draw[->] (l) to (dp);
\draw[->] (pstar) -- (dp);
\draw[->] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
\draw[->] (dp) -- (p2);
\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
\draw[->] (p2) to (pstar2);
\draw[->] (dp2) -- (y);
\draw[->] (cd) -- (dp2);
\draw[->] (cd) -- (y);
\draw[->] (l) -- (p2);
\draw[->] (l) to [out=0,in=90, looseness=0.4] (dp2);
\draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
\draw[->] (pstar2) -- (dp2);
\draw[->, semithick] (p2) to [out=+45,in=+135, looseness=0.3] (y);
\end{tikzpicture}
```

We now turn to an application where we show how these challenges translate into analytic decisions. We will show ways to mitigate or better understand bias to the best of the available data's abilities in an attempt to inform the possible effect of Pin1 on all cause-dementia.

### 2.2. The Rotterdam Study
